### 2018-07-30 chapters/02-background/related-work.tex
Put CDF citation to footnote and solve back-quote errors

The semantics must be well known before an expression can be translated. There are two main approaches to solve that problem: (1) someone could specify the semantic information during the writing process (pre-defined semantics); and (2) the translator can find out the right semantic information in general mathematical expressions before it translates the expression. So-called \textit{interactive documents}\footnote{There is no adequate definition what interactive documents are. However, this name is widely used to describe electronic document formats that allow for interactivity to change the content in real time.}, such as the \gls*{cdf}~\parencite{CDF:Wolfram} by Wolfram Research, or \textit{worksheets} by \Maple{}, try to solve this problem with the approach (2) and allow one to embed semantic information into the input. Those complex document formats require specialized tools to show and work with the documents (Wolfram CDF Player, or \Maple{} for the \textit{worksheets}). The \JOBAD{} architecture~\parencite{JOBAD:orig} is able to create web-based interactive documents and uses \gls*{omdoc}~\parencite{OMDoc} to carry semantics. The documents can be viewed and edited in the browser. Those \JOBAD{-documents} also allow one to perform computations via \gls*{cas}. This gives one the opportunity to calculate, compute and change mathematical expressions directly in the document. The translation performs in the background, invisible to the user. Similar to the \JOBAD{} architecture, other interactive web documents exist, such as \textit{MathDox}~\parencite{MathDox} and \textit{The Planetary System}~\parencite{Planetary}. All of these demonstrate the potential of the educational system.

>>>>

The semantics must be well known before an expression can be translated. There are two main approaches to solve that problem: (1) someone could specify the semantic information during the writing process (pre-defined semantics); and (2) the translator can find out the right semantic information in general mathematical expressions before it translates the expression. So-called \textit{interactive documents}\footnote{There is no adequate definition what interactive documents are. However, this name is widely used to describe electronic document formats that allow for interactivity to change the content in real time.}, such as the \gls*{cdf}\footnote{Wolfram Research; \textit{Computable Document Format} (CDF); \url{http://www.wolfram.com/cdf/}, July 2011} by Wolfram Research, or \textit{worksheets} by \Maple{}, try to solve this problem with the approach (2) and allow one to embed semantic information into the input. Those complex document formats require specialized tools to show and work with the documents (Wolfram CDF Player, or \Maple{} for the \textit{worksheets}). The \JOBAD{} architecture~\parencite{JOBAD:orig} is able to create web-based interactive documents and uses \gls*{omdoc}~\parencite{OMDoc} to carry semantics. The documents can be viewed and edited in the browser. Those \JOBAD{-documents} also allow one to perform computations via \gls*{cas}. This gives one the opportunity to calculate, compute and change mathematical expressions directly in the document. The translation performs in the background, invisible to the user. Similar to the \JOBAD{} architecture, other interactive web documents exist, such as \textit{MathDox}~\parencite{MathDox} and \textit{The Planetary System}~\parencite{Planetary}. All of these demonstrate the potential of the educational system.

### 2018-07-30 chapters/03-translator/forward-trans.tex
Put CDF citation to footnote and solve back-quote errors

\begin{figure}[ht]
	\centering
	\includegraphics[clip, trim=0.2cm 0.2cm 0.2cm 0.2cm, scale=0.75]{SyntaxTreeUseCase.pdf}
	\caption{pom-pt for a Jacobi polynomial using the DLMF/DRMF \LaTeX{} macro. Each leaf contains information from the lexicon files.}
	\label{fig:syntax-tree-usecase}
	\vspace{-10pt}
\end{figure}

@@

The \gls*{bnf} grammar defines some basic grammatical rules for generic \LaTeX{} macros, such as for \verb|\frac|, \verb|\sqrt|. Therefore, there is a hierarchical structure for those symbols similar to the structure in expression trees. As already mentioned, some of these types can be translated directly, such as Greek letters, while others are more complex, such as semantic \LaTeX{} macros. Therefore, the translators delegate the translation to specialized subtranslators. This delegation process is implemented in \Cref{line:transLeaf,line:transSeq} of Algorithm~{alg:translation}. The Subsection~\ref{sec:subtranslators} discusses these classes in more detail.

@@

In mathematical expressions, notations are mostly mixed, depending on the case and number of operands. For example, infix notation is common for binary operators ($+$, $-$, $\cdot$, $\mod$ etc.), while functional notations are conveniently used for any kind of functions ($sin$, $cos$, etc.). Sometimes the same symbol is used in different notations to distinguish different meanings. For example, the '$-$' as a unary operator is used in prefix notation to indicate the negative value of its operand, such as in '$-2$'. Of course, $-$ can also be the binary operator for subtraction, which is commonly used in infix notation. An example for the postfix notation is factorial, such as '$2!$'.

Most programming languages (and \gls*{cas} as well) internally use prefix or postfix notation and do not mix the notations in one expression, since it is more convenient to parse expressions in uniform notations. However, the common practice in science is to use mixed notations in expressions. Since the \gls*{pom} has rarely implemented mathematical grammatical rules yet, it takes the input as it is and does not build an expression tree. Therefore, it parses all four examples from~\Cref{tab:notations} to four different \gls*{pom-pt}s rather than to one unique expression tree. In general, this is not a problem for our translation process since most \gls*{cas} are familiar with the most common notations. Therefore, the translator does not need to know that $a$ and $b$ are the operands of the binary operator '$+$' in '$a+b$.' The translator could simply translate the symbols in '$a+b$' in the same order as they appear in the expression and the \gls*{cas} would understand it. However, this simple approach generates two problems.
\begin{enumerate}%[label=\arabic*)]
\item \label{prob:1} The translated expression is only syntactically correct if the input expression was syntactically correct.
\item \label{prob:2} We cannot translate expressions to a \gls*{cas} which uses a non-standard notation.
\end{enumerate}

@@

In most cases, parentheses can be used to disambiguate expressions. We used them in~\Cref{tab:amb_ex} to clarify the different interpretations in example 2. But sometimes, even parentheses cannot solve a mistaken computation. For example, there is no way to add parentheses to force \Maple{} to compute $n!!$ as the double factorial function. Even $(n)!!$ will be interpreted as $(n!)!$. Rather than using the exclamation mark in \Maple, one could also use the functional notation. For example, the interpretations $(2!)!$ and $(2)!!$ can be distinguished in \Maple{} by using \verb|factorial(factorial(2))| and \verb|doublefactorial(2)| respectively. We define the translations as follows:

@@

A \verb|SequenceTranslator| translates the \textit{sequence} and \textit{balanced expressions} in the \gls*{pom-pt}. If a node $n$ is a leaf and the represented symbol an open bracket (parentheses, square brackets and so on) the following nodes are also taken as a \textit{sequence}. Hence, combined with the recursive translation approach, the \verb|SequenceTranslator| also checks balances of parentheses in expressions. An expression such as '$(a]$' is producing a mismatched parentheses error. On the other hand, this is a problem for interval expressions such as '$[a,b)$'. In the current version, the program cannot distinguish between mismatched parentheses and half-opened, half-closed intervals. Whether an expression is an interval or another expression is difficult to decide and can depend on the context. Also, the parenthesis checker could simply be deactivated to allow mismatched parentheses in an expression. But this functionality is not implemented yet.

The \verb|SequenceTranslator| also handle positions of multiplication symbols. There are a couple of obvious choices to translate multiplication signs. The most common symbol for multiplications is still the white space or none symbol at all, as explained previously. Consider the simple expression $2n\pi$. The \gls*{pom-pt} generates a sequence node with three children, namely $2$, $n$ and $\pi$. This sequence should be interpreted as a multiplication of the three elements. The \verb|SequenceTranslator| checks the types of the current and next nodes in the tree to decide if it should add a multiplication symbol or not. For example, if the current or next node is an operator, a relation symbol or an ellipsis, there will be no multiplication symbol added. However, this approach implies an important property. The translator interprets all sequences of nodes as multiplications as long as it is not defined otherwise. This potentially produces strange effects. Consider an expression such as $f(x)$. Translate this to \Maple{} will be \verb|f*(x)|. But we do not consider this translation to be wrong, because there is a semantic macro to represent functions. In this case, the user should use \verb|\f{f}@{x}| instead of \verb|f(x)| to distinguish between $f$ as a function call and $f$ as a symbol.

>>>>

\begin{figure}[ht]
	\centering
	\includegraphics[clip, trim=0.2cm 0.2cm 0.2cm 0.2cm, scale=0.75]{SyntaxTreeUseCase.pdf}
	\caption{The \gls*{pom-pt} for the Jacobi polynomial example~\eqref{eq:P} using the DLMF/DRMF \LaTeX{} macro. Each leaf contains information from the lexicon files.}
	\label{fig:syntax-tree-usecase}
	\vspace{-10pt}
\end{figure}

@@

The \gls*{bnf} grammar defines some basic grammatical rules for generic \LaTeX{} macros, such as for \verb|\frac|, \verb|\sqrt|. Therefore, there is a hierarchical structure for those symbols similar to the structure in expression trees. As already mentioned, some of these types can be translated directly, such as Greek letters, while others are more complex, such as semantic \LaTeX{} macros. Therefore, the translators delegate the translation to specialized subtranslators. This delegation process is implemented in \Cref{line:transLeaf,line:transSeq} of Algorithm~\ref{alg:translation}. The Subsection~\ref{sec:subtranslators} discusses these classes in more detail.

@@

In mathematical expressions, notations are mostly mixed, depending on the case and number of operands. For example, infix notation is common for binary operators ($+$, $-$, $\cdot$, $\mod$, etc.), while functional notations are conveniently used for any kind of functions ($sin$, $cos$, etc.). Sometimes the same symbol is used in different notations to distinguish different meanings. For example, the `$-$' as a unary operator is used in prefix notation to indicate the negative value of its operand, such as in `$-2$'. Of course, `$-$' can also be the binary operator for subtraction, which is commonly used in infix notation. An example for the postfix notation is factorial, such as `$2!$'.

Most programming languages (and \gls*{cas} as well) internally use prefix or postfix notation and do not mix the notations in one expression, since it is more convenient to parse expressions in uniform notations. However, the common practice in science is to use mixed notations in expressions. Since the \gls*{pom} has rarely implemented mathematical grammatical rules yet, it takes the input as it is and does not build an expression tree. Therefore, it parses all four examples from~\Cref{tab:notations} to four different \gls*{pom-pt}s rather than to one unique expression tree. In general, this is not a problem for our translation process since most \gls*{cas} are familiar with the most common notations. Therefore, the translator does not need to know that `$a$' and `$b$' are the operands of the binary operator `$+$' in `$a+b$.' The translator could simply translate the symbols in `$a+b$' in the same order as they appear in the expression and the \gls*{cas} would understand it. However, this simple approach generates two problems.
\begin{enumerate}%[label=\arabic*)]
\item \label{prob:1} The translated expression is only syntactically correct if the input expression was syntactically correct.
\item \label{prob:2} We cannot translate expressions to a \gls*{cas} which uses a non-standard notation.
\end{enumerate}

@@

In most cases, parentheses can be used to disambiguate expressions. We used them in~\Cref{tab:amb_ex} to clarify the different interpretations in example 2. But sometimes, even parentheses cannot solve a mistaken computation. For example, there is no way to add parentheses to force \Maple{} to compute `$n!!$' as the double factorial function. Even `$(n)!!$' will be interpreted as `$(n!)!$'. Rather than using the exclamation mark in \Maple, one could also use the functional notation. For example, the interpretations `$(2!)!$' and `$(2)!!$' can be distinguished in \Maple{} by using \verb|factorial(factorial(2))| and \verb|doublefactorial(2)| respectively. We define the translations as follows:

@@

A \verb|SequenceTranslator| translates the \textit{sequence} and \textit{balanced expressions} in the \gls*{pom-pt}. If a node $n$ is a leaf and the represented symbol an open bracket (parentheses, square brackets and so on) the following nodes are also taken as a \textit{sequence}. Hence, combined with the recursive translation approach, the \verb|SequenceTranslator| also checks balances of parentheses in expressions. An expression such as `$(a]$' is producing a mismatched parentheses error. On the other hand, this is a problem for interval expressions such as `$[a,b)$'. In the current version, the program cannot distinguish between mismatched parentheses and half-opened, half-closed intervals. Whether an expression is an interval or another expression is difficult to decide and can depend on the context. Also, the parenthesis checker could simply be deactivated to allow mismatched parentheses in an expression. But this functionality is not implemented yet.

The \verb|SequenceTranslator| also handle positions of multiplication symbols. There are a couple of obvious choices to translate multiplication signs. The most common symbol for multiplications is still the white space or none symbol at all, as explained previously. Consider the simple expression `$2n\pi$'. The \gls*{pom-pt} generates a sequence node with three children, namely $2$, $n$ and $\pi$. This sequence should be interpreted as a multiplication of the three elements. The \verb|SequenceTranslator| checks the types of the current and next nodes in the tree to decide if it should add a multiplication symbol or not. For example, if the current or next node is an operator, a relation symbol or an ellipsis, there will be no multiplication symbol added. However, this approach implies an important property. The translator interprets all sequences of nodes as multiplications as long as it is not defined otherwise. This potentially produces strange effects. Consider an expression such as `$f(x)$'. Translate this to \Maple{} will be \verb|f*(x)|. But we do not consider this translation to be wrong, because there is a semantic macro to represent functions. In this case, the user should use \verb|\f{f}@{x}| instead of \verb|f(x)| to distinguish between $f$ as a function call and $f$ as a symbol.
